Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils

### 2020-10-13 doc/manuscript.Rmd
minor ms corrections

Ecosystems adapt differently to global warming through microclimate factors. Mires are sensitive wetland habitats that strongly rely on local soil properties, which makes them a good model to understand how local climate parameters counteract the effects of climate change. We quantified the temperature buffering effect in waterlogged mire soils as compared with adjacent dry soils.

@@

Waterlogged soils had less diurnal thermal amplitude (-2.3 ºC), less annual thermal amplitude (-5.1 ºC), cooler summer maximums (-4.3 ºC) and warmer winter minimums (+0.8 ºC). CHELSA air temperatures only correlated significantly (p < 0.05) with the winter minimum temperatures (Pearson's r > 0.83), and CHELSA predictions were less accurate (higher RMSE) for waterlogged soils, except for the summer maximums.

@@

Since root-zone temperature is a major determinant of plant ecophysiology [@RN3024], the groundwater buffer effect is expected to allow mire plants to live in a wider range of air temperatures than they could otherwise. Indeed, mires have a relatively homogeneous flora across their range despite their broad distribution. In Europe, fens are distributed from the Iberian Peninsula to boreal Fennoscandia and from low valleys to the alpine belt [@RN4678]. In the arctic, temperate-continental and suboceanic regions of boreal Russia mires make up an important part of the landscape, from 20 to 80 % of the surface of different regions stretching from Europe to the Pacific [@RN4937]. In the United States, fens are distributed across the glaciated Midwest and Northeast, as well as portions of the Appalachian Mountains and mountainous West [@RN4933]. However, continued aridification is expected to significantly reduce the overall extent of wetlands as it has been shown in the Midwestern United States [@RN4934].

It is evident that the groundwater buffer effect will play a determinant role in the response of mire habitats to climate change [@RN4936]. Both bogs and fens are actively peat forming and are dependent on precipitation, whereas fens rely also on sources of telluric water from mineral ground. Moreover, they show a mosaic of microreliefs and a specific plant cover zonation which makes them especially sensitive to small changes in wetness. Climate change affects the quantity, timing and spatial distribution of precipitation, leading to changes in surface wetness which alter the intensity or organic decomposition by disturbing the conditions for plant grow and the depth of air penetration [@RN4935]. Global warming would also result in warmer groundwater delivered to fens, but there is a gap of knowledge about the rate of such changes and their ecological consequences within the mires [@RN4936].

@@

Most models currently employed to predict vegetation and species distribution use macroclimatic parameters, like the ones obtained from CHELSA [@RN4766]. This study revealed that CHELSA values correlate relatively well with soil temperatures in the case of the minimum temperatures of the coldest month, but not with the rest of the bioclimatic variables under examination. Moreover, CHELSA predictions are less accurate for waterlogged soils, except for the summer max, which might be the consequence of the noise introduced by sun-heated soil. In addition to the water buffering effect, other features may explain the differences between our results and CHELSEA values, like the height of temperature measurement (5 cm below the upper layer of the soil in our study versus 2 m above the ground in CHELSA); the temporal scale of observation (hourly based for a 5 years period in our study versus a 34 years period); and the spatial resolution (c. 1 m in our study versus c. 1 km). Be that as it may, our results bring out the importance of considering local factors as drivers of microclimatic changes [@RN4762]. These factors contribute to landscape heterogeneity, producing safe sites that can act as micro-refuges and buffer species from regional climatic warming.

>>>>

Ecosystems adapt differently to global warming through microclimatic factors. Mires are sensitive wetland habitats that strongly rely on local soil properties, which makes them a good model to understand how local climatic parameters counteract the effects of climate change. We quantified the temperature buffering effect in waterlogged mire soils as compared with adjacent dry soils.

@@

Waterlogged soils had less diurnal thermal amplitude (-2.3 ºC), less annual thermal amplitude (-5.1 ºC), cooler summer maximums (-4.3 ºC) and warmer winter minimums (+0.8 ºC). CHELSA air temperatures only correlated significantly (p < 0.05) with winter minimum soil temperatures (Pearson's r > 0.83), and CHELSA predictions were less accurate (higher RMSE) for waterlogged soils, except for the summer maximums.

@@

Since root-zone temperature is a major determinant of plant ecophysiology [@RN3024], the groundwater buffer effect is expected to allow mire plants to live in a wider range of air temperatures than they could otherwise. Indeed, mires have a relatively homogeneous flora across their range, despite their broad distribution. In Europe, fens are distributed from the Iberian Peninsula to boreal Fennoscandia and from low valleys to the alpine belt [@RN4678]. In the arctic, temperate-continental and suboceanic regions of boreal Russia mires make up an important part of the landscape, from 20 to 80 % of the surface of different regions stretching from Europe to the Pacific [@RN4937]. In the United States, fens are distributed across the glaciated Midwest and Northeast, as well as portions of the Appalachian Mountains and mountainous West [@RN4933]. However, continued aridification is expected to significantly reduce the overall extent of wetlands as it has been shown in the Midwestern United States [@RN4934].

It is evident that the groundwater buffer effect will play a determinant role in the response of mire habitats to climate change [@RN4936]. Both bogs and fens are actively peat forming and depend on precipitation, but fens rely also on sources of telluric water from mineral ground. Moreover, they show a mosaic of microreliefs and a specific plant cover zonation which makes them especially sensitive to small changes in wetness. Climate change affects the quantity, timing and spatial distribution of precipitation, leading to changes in surface wetness which alter the intensity or organic decomposition by disturbing the conditions for plant grow and the depth of air penetration [@RN4935]. Global warming would also result in warmer groundwater delivered to fens, but there is a gap of knowledge about the rate of such changes and their ecological consequences within the mires [@RN4936].

@@

Most models currently employed to predict vegetation and species distribution use macroclimatic parameters, like the ones obtained from CHELSA [@RN4766]. This study revealed that CHELSA values correlate relatively well with soil temperatures in the case of the minimum temperatures of the coldest month, but not with the rest of the bioclimatic variables under examination. Moreover, CHELSA predictions are less accurate for waterlogged soils, except for the summer max, which might be the consequence of the noise introduced by sun-heated soil. In addition to the water buffering effect, other features may explain the differences between our results and CHELSEA values, like the height of temperature measurement (5 cm below the upper layer of the soil in our study versus 2 m above the ground in CHELSA); the temporal scale of observation (hourly based for a 5-year period in our study versus a 34-year period); and the spatial resolution (c. 1 m in our study versus c. 1 km). Be that as it may, our results bring out the importance of considering local factors as drivers of microclimatic changes [@RN4762]. These factors contribute to landscape heterogeneity, producing safe sites that can act as micro-refuges and buffer species from regional climatic warming.
